---
title: "Indirect Cost Forecasting"
author: "Md Ismail Hossain"
date: "2025-01-30"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE)
```

\newpage

# Introduction
This report presents the analysis and forecasting of indirect costs using three different models: **SARIMA, ETS, and Prophet**. The objective is to identify the most suitable forecasting method and predict future costs with high accuracy.

# Understanding the Data
The dataset consists of **monthly indirect cost data from July 2021 to January 2025**. A preliminary analysis revealed **seasonality and trend variations**, necessitating an advanced forecasting approach.


```{r}
# Load necessary libraries
library(forecast)
library(tseries)
library(readr)
library(readxl)
library(ggplot2)
library(prophet)


# Read the data
data <- read_excel("C:/Users/mhossa11/OneDrive - University of Wyoming/Projects/Indirect Cost Forecasting 02022025/Data/Monthly IC Expenditures FY22-partial FY25.xlsx", 
    sheet = "Clean Data")

# Convert Date column to Date format
data$Date <- as.Date(data$Date, format="%m/%d/%Y")

# Create a time series object
ts_data <- ts(data$Indirect_Cost, start=c(2021,7), frequency=12)  # Monthly data starting from July 2021

# Create ggplot visualization
ggplot(data, aes(x = Date, y = Indirect_Cost)) +
  geom_line(color = "blue", size = 1) +  # Line color and thickness
  geom_point(color = "red", size = 2) +  # Points for better visibility
  scale_x_date(date_breaks = "3 months", date_labels = "%b-%Y") +  # Show x-axis labels for every 3 months
  labs(title = "Indirect Cost Over Time",
       x = "Date",
       y = "Indirect Cost") +
  theme_minimal() +  # Apply a minimal clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis labels
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))



```




# Selecting Best Forecasting Model



We compared three forecasting models:

 - **SARIMA (Seasonal ARIMA)**
 - **ETS (Exponential Smoothing)**
 - **Prophet (Additive Time Series Model by Facebook)**

Each model was trained on **80% of the dataset** and evaluated using **Root Mean Squared Error (RMSE)**. The results were:

\[
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}
\]

\begin{table}[h]
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Model} & \textbf{RMSE (Lower is Better)} \\
\hline
SARIMA & 27,8408.8 \\
Prophet & \textbf{26,3486.9} \\
ETS & 37,5984.8 \\
\hline
\end{tabular}
\caption{Model Performance Comparison Based on RMSE}
\end{table}

The **ETS model performed the worst**, while **Prophet had the lowest RMSE, making it the best model** for forecasting indirect costs.

\textbf{Conclusion:} Prophet was selected as the best model due to its superior performance in capturing seasonality and trends.




```{r}
# --------- 1️⃣ TRAIN-TEST SPLIT ---------
train_size <- round(length(ts_data) * 0.8)
train_data <- ts_data[1:train_size]
test_data <- ts_data[(train_size+1):length(ts_data)]

# --------- 2️⃣ SARIMA MODEL ---------
# Fit SARIMA Model (d=0 since ADF test confirmed stationarity)
sarima_model <- auto.arima(train_data, seasonal=TRUE, d=0)

# Forecast using SARIMA
sarima_forecast <- forecast(sarima_model, h=length(test_data))

# Compute RMSE for SARIMA
rmse_sarima <- sqrt(mean((sarima_forecast$mean - test_data)^2))
#print(paste("RMSE for SARIMA:", rmse_sarima))

# --------- 3️⃣ PROPHET MODEL ---------
# Prepare data for Prophet (Prophet requires specific column names)
prophet_df <- data.frame(ds=data$Date[1:train_size], y=train_data)

# Fit the Prophet model
prophet_model <- prophet(prophet_df)

# Create future dataframe for Prophet
future <- make_future_dataframe(prophet_model, periods=length(test_data), freq="month")

# Forecast using Prophet
prophet_forecast <- predict(prophet_model, future)

# Extract the forecast for test period
prophet_predictions <- prophet_forecast$yhat[(nrow(prophet_forecast)-length(test_data)+1):nrow(prophet_forecast)]

# Compute RMSE for Prophet
rmse_prophet <- sqrt(mean((prophet_predictions - test_data)^2))
#print(paste("RMSE for Prophet:", rmse_prophet))

# --------- 4️⃣ ETS MODEL ---------
# Fit ETS Model
ets_model <- ets(train_data)

# Forecast using ETS
ets_forecast <- forecast(ets_model, h=length(test_data))

# Compute RMSE for ETS
rmse_ets <- sqrt(mean((ets_forecast$mean - test_data)^2))
#print(paste("RMSE for ETS:", rmse_ets))

# --------- 5️⃣ VISUALIZATION ---------
# Combine forecast results for comparison
forecast_df <- data.frame(
  Date = data$Date[(train_size+1):length(ts_data)],
  Actual = test_data,
  SARIMA = sarima_forecast$mean,
  Prophet = prophet_predictions,
  ETS = ets_forecast$mean
)

# Plot Actual vs Predicted for SARIMA, Prophet, and ETS
ggplot(forecast_df, aes(x=Date)) +
  geom_line(aes(y=Actual, color="Actual"), size=1.2) +
  geom_line(aes(y=SARIMA, color="SARIMA Forecast"), size=1) +
  geom_line(aes(y=Prophet, color="Prophet Forecast"), size=1) +
  geom_line(aes(y=ETS, color="ETS Forecast"), size=1) +
  labs(title="SARIMA vs. Prophet vs. ETS Forecasting Comparison",
       x="Date", y="Indirect Cost") +
  scale_color_manual(name="Legend",
                     values=c("Actual"="black", "SARIMA Forecast"="blue", 
                              "Prophet Forecast"="green", "ETS Forecast"="red")) +
  theme_minimal()

# --------- 6️⃣ PRINT COMPARISON RESULTS ---------
print("RMSE Comparison:")
rmse_results <- data.frame(Model = c("SARIMA", "Prophet", "ETS"),
                           RMSE = c(rmse_sarima, rmse_prophet, rmse_ets))
#print(rmse_results)

# Find the best model
best_model <- rmse_results$Model[which.min(rmse_results$RMSE)]
print(paste("Best Model Based on RMSE:", best_model))
```

# Using Prophet Model to Forcast

Since Prophet performed best, we trained it on the **entire dataset** and generated forecasts for the next **12 months**.


```{r}
# Load necessary libraries
library(prophet)
library(ggplot2)

# Read the data
data <- read_excel("C:/Users/mhossa11/OneDrive - University of Wyoming/Projects/Indirect Cost Forecasting 02022025/Data/Monthly IC Expenditures FY22-partial FY25.xlsx", 
    sheet = "Clean Data")

# Convert Date column to Date format
data$Date <- as.Date(data$Date, format="%m/%d/%Y")

# Rename columns for Prophet
df <- data.frame(ds = data$Date, y = data$Indirect_Cost)

# --------- 1️⃣ FIT PROPHET MODEL ---------
# Initialize and fit Prophet model
prophet_model <- prophet(df,interval.width = 0.95)

# --------- 2️⃣ FORECAST NEXT 12 MONTHS ---------
# Create future dates (12 months ahead)
future <- make_future_dataframe(prophet_model, periods = 12, freq = "month")

# Predict future values
forecast <- predict(prophet_model, future)


# --------- 4️⃣ PLOT COMPONENTS (Trend & Seasonality) ---------
#prophet_plot_components(prophet_model, forecast)



```


# Next 12 Month Forecast

\textbf{Visualization:} The Prophet model effectively captures seasonal patterns and upward trends.

```{r}
# --------- 3️⃣ PLOT THE FORECAST ---------
# Plot forecasted values
plot(prophet_model, forecast) +
  ggtitle("Forecast for Indirect Cost") +
  xlab("Date") +  # Set X-axis label
  ylab("Indirect Cost")  # Set Y-axis label
```


```{r}
# --------- 5️⃣ PRINT FORECASTED VALUES ---------
# Display forecasted data (next 12 months)
forecasted_value <- tail(forecast[, c("ds", "yhat", "yhat_lower", "yhat_upper")],12)
colnames(forecasted_value) <- c("Date", "Forecast", "Lower CI (95%)", "Upper CI (95%)")
forecasted_value

write.csv(forecasted_value,"forecasted_value_feb_25_to_Jan_26.csv")
```

## Interpretation of Forecast
- A **steady increase in indirect costs** is projected.
- Costs are expected to exceed **$2,100,000 by June 2025**.
- The **confidence interval indicates possible fluctuations**. The 95% confidence interval (CI) for the forecasted values tells us the range in which we expect the actual indirect costs to fall, 95% of the time.

Think of it as a margin of error around our prediction.






# Conclusion and Recommendations

## Conclusion
- The **Prophet model outperformed SARIMA and ETS**, achieving the lowest RMSE.
- The forecast suggests **a steady rise in costs**, emphasizing the need for financial planning.
- Confidence intervals indicate **potential variations**, requiring periodic monitoring.

## Recommendations
1. **Monitor monthly actual costs** and compare with forecasts for adjustments.
2. **Retrain the model periodically** to incorporate new data.
