---
title: 'Indirect Cost Forecasting: Regression Approach'
author: "Md Ismail Hossain"
date: "2025-03-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

\newpage

# Data Structure:

## Independent Variables:

 - Award Number [Award External Funding By Dept_Details.xlsx]
 - Award Fund Amount [Award External Funding By Dept_Details.xlsx]
 - Indirect Cost Rate [Column Burden Schedule in Project Committed Cost Details_Results.xlsx]: From F and A Schedule Descriptions_Results.xlsx try to understand the explanaiton of  F&A Schedule and convert that Burden Schedule to a numberic value to multiply with the Award Fund Amount to calculate the Indirect cost amount.
 - Grant Funding Type [Project Financial Summary_Results.xlsx]
 - Grant Duration [Award Start Date - Award End Date] [Award Personnel Details_Current Awards.xlsx]
 
 
 
 Project Information_Results.xlsx data have:
 - Award Number
 - Award Type
 - Project Funding Type
 - Award Start Date
 - Award End Date
 - Award F&A Schedule
 
 
 
 
 
## Dependent Variables:

 - Length of spending [First Spending Date - Award Date]: Either we have to use the first Accounting Date or Expenditure Date as the First Spending Date from Project Committed Cost Details_Results.xlsx data. Or should I use Expenditure Date from Project Expenditure Details_Results.xlsx data???


# Coding:


```{r}
library(readxl)
library(dplyr)
library(lubridate)

# Load datasets
award_funding <- read_excel("/mnt/data/Award External Funding By Dept_Details.xlsx")
personnel_awards <- read_excel("/mnt/data/Award Personnel Details_Current Awards.xlsx")
fa_schedule <- read_excel("/mnt/data/F and A Schedule Descriptions_Results.xlsx")
committed_costs <- read_excel("/mnt/data/Project Committed Cost Details_Results.xlsx")
expenditure_details <- read_excel("/mnt/data/Project Expenditure Details_Results.xlsx")
financial_summary <- read_excel("/mnt/data/Project Financial Summary_Results.xlsx")
project_info <- read_excel("/mnt/data/Project Information_Results.xlsx")

date_conversion <- function(date_column) {
  as.Date(date_column, format = "%Y-%m-%d")
}

# Extract Indirect Cost Rate from FA Schedule
tidy_fa_schedule <- fa_schedule %>%
  select(`Burden Schedule`, `F&A Rate`) %>%
  mutate(`F&A Rate` = as.numeric(`F&A Rate`))

# Merge Award Data with Project Information
merged_data <- project_info %>%
  left_join(award_funding, by = "Award Number") %>%
  left_join(tidy_fa_schedule, by = c("Award F&A Schedule" = "Burden Schedule")) %>%
  mutate(
    Indirect_Cost_Amount = `Award Fund Amount` * `F&A Rate`,
    Award_Start_Date = date_conversion(`Award Start Date`),
    Award_End_Date = date_conversion(`Award End Date`),
    Grant_Duration = as.numeric(Award_End_Date - Award_Start_Date)
  )

# Determine First Spending Date
first_spending_date <- committed_costs %>%
  group_by(`Award Number`) %>%
  summarise(First_Spending_Date = min(date_conversion(`Accounting Date`)))

# Merge with expenditure data if necessary
first_spending_date_exp <- expenditure_details %>%
  group_by(`Award Number`) %>%
  summarise(First_Expenditure_Date = min(date_conversion(`Expenditure Date`)))

# Choose the earliest spending date
spending_dates <- first_spending_date %>%
  left_join(first_spending_date_exp, by = "Award Number") %>%
  mutate(First_Spending_Date = pmin(First_Spending_Date, First_Expenditure_Date, na.rm = TRUE))

# Merge with main dataset
final_data <- merged_data %>%
  left_join(spending_dates, by = "Award Number") %>%
  mutate(
    Length_of_Spending = as.numeric(First_Spending_Date - Award_Start_Date)
  )

# Select relevant columns
final_dataset <- final_data %>%
  select(Award_Number, `Award Fund Amount`, Indirect_Cost_Amount, Grant_Duration, Length_of_Spending, `Grant Funding Type`)

# Save the final dataset to a CSV file
write.csv(final_dataset, "/mnt/data/final_dataset.csv", row.names = FALSE)

print("Final dataset created successfully.")

```

